The Iterative Signature Algorithm for Gene Expression data

### 2009-07-14 eisa/inst/doc/EISA_tutorial.Rnw
Some eisa vignette improvements.

@@

For extracting the results themselves, the \Rfunction{summary}
function can be used, this converts them to a simple data frame. The
$p$-value limit can be supplied to \Rfunction{summary}. Note, that
since \Rfunction{isa.GO} calculates enrichment for many gene sets
(i.e. for all biclusters), \Rfunction{summary} returns a list of data
frames, one for each bicluster. A table for the first module:
<<GO summary>>=
summary(GO[[1]], p=0.001)[[1]]
@

The \texttt{drive} column includes the Entrez ids of the genes that
are included in the intersection of the bicluster and the GO
category. They can be also obtained with the
\Rfunction{geneIdsByCategory} function. The following returns the
genes in the first module and the second GO BP category.
<<GO gene ids by category>>=
geneIdsByCategory(GO[[1]])[[1]][[2]]
@

You can use the \Rpackage{GO.db} package to obtain more information
about the enriched GO categories.
<<GO info>>=
sigCategories(GO[[1]])[[1]]
library(GO.db)
mget(sigCategories(GO[[1]])[[1]], GOTERM)
@

@@

Enrichment calculation against the KEGG pathway goes essentially the
same way as for the Gene Ontology, this time we use the
\Rfunction{isa.KEGG} function:
<<KEGG enrichment>>=
KEGG <- isa.KEGG(modules)
KEGG
summary(KEGG)[[1]]
library(KEGG.db)
mget(sigCategories(KEGG)[[1]], KEGGPATHID2NAME)
@

@@

The \Rpackage{eisa} includes a simple way to check whether the genes
in a bicluster are associated with a chromosome. See the
\Rfunction{isa.CHR} function:
<<CHR enrichment>>=
CHR <- isa.CHR(modules)
summary(CHR,p=0.05)[[11]]
@

The third bicluster has \Sexpr{summary(CHR,p=0.05)[[2]]$Count[1]} % $
genes on chromosome \Sexpr{rownames(summary(CHR,p=0.05)[[2]])[1]}.

@@

The enrichment calculation itself is basically the same as for GO and
KEGG, but the \Rfunction{isa.miRNA} function should be used:
<<miRNA enrichment>>=
if (require(targetscan.Hs.eg.db)) {
  miRNA <- isa.miRNA(modules)
  summary(miRNA, p=1)[[7]]
}
@

@@

The \Rpackage{biclust} R package implements several biclustering
algorithms in a unified framework. It uses the class \Rclass{Biclust}
to store a set of biclusters. The \Rfunction{eisa.biclust} converts ISA
modules to a \Rclass{Biclust} object. This requires the binarization
of the modules, i.e. the ISA scores are lost, they are converted to
zeros and ones:
<<biclust,cache=TRUE>>=
library(biclust)
Bc <- eisa.biclust(modules)
Bc
@

@@

You can also use the \Rpackage{biclust} package to create image
plots, by calling the \Rfunction{drawHeatmap} function. The result is
in Fig.~\ref{fig:drawHeatmap}.
<<heatmap biclust pre,echo=FALSE,results=hide>>=
CairoPNG("heatmap.png", 400, 400 )
@ 
<<heatmap>>=
drawHeatmap(exprs(ALL), Bc, number=1) 
@
<<heatmap post,echo=FALSE,results=hide>>=
dev.off() 
@ 
\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{heatmap}
  \caption{TODO}
  \label{fig:drawHeatmap}
\end{figure}

@@

Profile plots visualize the difference between the genes (or samples)
in the modules and the rest of the expression data. A profile plot
contains a line plot for every single gene (or sample) and the genes
that belong to the module have a different color, see
Fig.~\ref{fig:parallel}.
<<parallel.pre,echo=FALSE,results=hide>>=
CairoPNG("parallel.png", 400, 400 )
@ 
<<parallel>>=
parallelCoordinates(exprs(ALL), Bc, number=1, plotBoth=TRUE, pch=NA)
@
<<parallel.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{parallel}
\caption{Profile plots for the first bicluster found by ISA.}
\label{fig:parallel}
\end{figure}

@@

\Rfunction{goplot} uses the \Rpackage{igraph} package to create a
graph with associated meta data:
<<gograph object>>=
summary(goplot.2)
@ 
The \texttt{width} and \texttt{height} graph attributes contain the
suggested width and height of the graph, if plotted to a bitmap
device. (The graph attributes of an \Rpackage{igraph} graph can be
queried with the `\texttt{\$}' selector:
<<gograph object 2>>=
goplot.2$width
goplot.2$height
@

@@

>>>>

<<echo=FALSE,print=FALSE>>=
options(width=60)
@

@@

For extracting the results themselves, the \Rfunction{summary}
function can be used, this converts them to a simple data frame. The
$p$-value limit can be supplied to \Rfunction{summary}. Note, that
since \Rfunction{isa.GO} calculates enrichment for many gene sets
(i.e. for all biclusters), \Rfunction{summary} returns a list of data
frames, one for each bicluster. A table for the first module:
<<GO summary>>=
summary(GO[[1]], p=0.001)[[2]][,-6]
@

We omitted the sixth column of the result, because it is very wide and
would look bad in this vignette. This column is called
\variable{drive} and lists the Entrez IDs of the genes that are in the
intersection of the bicluster and the GO category; or in other
words, the genes that drive the enrichment.
These genes can also be obtained with the
\Rfunction{geneIdsByCategory} function. The following returns the
genes in the second module and the third GO BP category. (The GO
categories are ordered according to the enrichment $p$-values, just
like in the output of \Rfunction{summary}.)
<<GO gene ids by category>>=
geneIdsByCategory(GO[[1]])[[2]][[3]]
@

You can use the \Rpackage{GO.db} package to obtain more information
about the enriched GO categories.
<<GO info>>=
sigCategories(GO[[1]])[[2]]
library(GO.db)
mget(sigCategories(GO[[1]])[[2]], GOTERM)
@

@@

Enrichment calculation against the KEGG pathway goes essentially the
same way as for the Gene Ontology, this time we use the
\Rfunction{isa.KEGG} function:
<<KEGG enrichment>>=
KEGG <- isa.KEGG(modules)
KEGG
summary(KEGG)[[4]]
library(KEGG.db)
mget(sigCategories(KEGG)[[4]], KEGGPATHID2NAME)
@

@@

The \Rpackage{eisa} includes a simple way to check whether the genes
in a bicluster are associated with a chromosome. See the
\Rfunction{isa.CHR} function:
<<CHR enrichment>>=
CHR <- isa.CHR(modules)
summary(CHR,p=0.05)[[11]][,-6]
@

The eleventh bicluster has \Sexpr{summary(CHR,p=0.05)[[11]]$Count[1]} % $
genes on chromosome
\Sexpr{rownames(summary(CHR,p=0.05)[[11]])[1]}. Here is a list of the
genes:
<<CHR enrichment list>>=
unlist(mget(geneIdsByCategory(CHR)[[11]][[1]], org.Hs.egSYMBOL))
@

@@

The enrichment calculation itself is basically the same as for GO and
KEGG, but the \Rfunction{isa.miRNA} function should be used:
<<miRNA enrichment>>=
if (require(targetscan.Hs.eg.db)) {
  miRNA <- isa.miRNA(modules)
  summary(miRNA, p=0.05)[[11]]
}
@

@@

The \Rpackage{biclust} R package implements several biclustering
algorithms in a unified framework. It uses the class \Rclass{Biclust}
to store a set of biclusters. The \Rfunction{eisa.biclust} function converts ISA
modules to a \Rclass{Biclust} object. This requires the binarization
of the modules, i.e. the ISA scores are lost, they are converted to
zeros and ones:
<<biclust,cache=TRUE>>=
library(biclust)
Bc <- eisa.biclust(modules)
Bc
@

@@

You can also use the \Rpackage{biclust} package to create image
plots, by calling the \Rfunction{drawHeatmap} function. The result is
in Fig.~\ref{fig:drawHeatmap}.
<<heatmap biclust pre,echo=FALSE,results=hide>>=
CairoPNG("heatmap.png", 400, 400 )
@ 
<<heatmap>>=
drawHeatmap(exprs(ALL[featureNames(modules),]), Bc, number=1) 
@
<<heatmap post,echo=FALSE,results=hide>>=
dev.off() 
@ 
\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{heatmap}
  \caption{TODO}
  \label{fig:drawHeatmap}
\end{figure}

@@

Profile plots visualize the difference between the genes (or samples)
in the modules and the rest of the expression data. A profile plot
contains a line plot for every single gene (or sample) and the genes
that belong to the module have a different color, see
Fig.~\ref{fig:parallel}.
<<parallel.pre,echo=FALSE,results=hide>>=
CairoPNG("parallel.png", 400, 400 )
@ 
<<parallel>>=
ex <- samp.exprs(isa.normalize(ALL[featureNames(modules),]))
parallelCoordinates(ex, Bc, number=2, pch=NA)
@
<<parallel.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{parallel}
\caption{Profile plots for the first bicluster found by ISA.}
\label{fig:parallel}
\end{figure}

@@

\Rfunction{goplot} uses the \Rpackage{igraph} package to create a
graph with associated meta data:
<<gograph object>>=
summary(goplot.2)
list.vertex.attributes(goplot.2)
@ 
The \texttt{width} and \texttt{height} graph attributes contain the
suggested width and height of the graph, if plotted to a bitmap
device. (The graph attributes of an \Rpackage{igraph} graph can be
queried with the `\texttt{\$}' selector:
<<gograph object 2>>=
goplot.2$width
goplot.2$height
@

@@

The \Rfunction{autogen.table} function creates a summary
